MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-26 main/MSstatsPTM_Main.tex
Implementation

The proposed statistical framework enables sample size calculation for future PTM experiments. 
The procedure has been described in general in~\cite{kutner_etal_04a}, and specifically for protein significance analysis in~\cite{oberg_vitek_09a}. 
It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively. Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow. Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$: \todo{There should be no hats in equations 11-13. The calculations assume that the variances are given externally, and are known. The input quantities above have no hats}
\begin{equation}
\widehat{SE}(\hat{\Delta}_{adj})^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{PTM}^{2} + \hat{\sigma}_{protein}^{2} \right) \right]
\leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Student distribution with $df = I(J-1)$ degrees of freedom.  Solving for $J$, the number of biological replicates per condition is
\begin{equation}
J \geq \frac{(2\hat{\sigma}_{PTM}^{2} + 2\hat{\sigma}_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
\end{equation}
\todo{Comment on the fact that since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation.} Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec. \ref{supp-sec:complex_design}}.

@@

@@

$MSstatsPTM$ requires the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e. the modification site). This will generally include the amino acid abbreviation, plus it's number in the protein sequence. For example, a modification on a serine, which was the 70th amino acid in the sequence would be marked ``S70''.

Occasionally tools will only include the peptide sequence with the modified amino acid highlighted. If there is no way to determine where in the protein sequence the modification occurred, $MSstatsPTM$ includes functionality to do this manually, given the modified peptide sequence and a FASTA file with the entire protein sequence. The package includes converters for seamless integration with multiple such tools. The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites. If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect. Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest. For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.

>>>>

The proposed statistical framework enables sample size calculation for future PTM experiments. 
The procedure has been described in general in~\cite{kutner_etal_04a}, and specifically for protein significance analysis in~\cite{oberg_vitek_09a}. 
It requires us to specify the desired levels of the following quantities: a) $q$, the False Discovery Rate of detecting differential abundance, b) $\beta$, the average Type II error rate, c) $\Delta_{adj}$, the minimal log$_2$-fold change in adjusted PTM abundance  of interest, d) $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison, and e) $\sigma_{PTM}^{2}$ and $\sigma_{protein}^{2}$, the anticipated variances associated with the summaries of the modified and unmodified features, respectively. Typically, the variances are estimated from an existing experiment, conducted with the same biological material and measurement workflow. Given the above quantities, and assuming a balanced design, the minimal number of replicates $J$ across $I$ conditions is chosen to bound the variance of the estimated log$_2$-fold change $\mathrm{SE}^{2}(\hat{\Delta}_{adj})$: \todo{There should be no hats in equations 11-13. The calculations assume that the variances are given externally, and are known. The input quantities above have no hats}
\begin{equation}
\widehat{SE}(\hat{\Delta}_{adj})^2 = \left[ \frac{2}{J} \left( \hat{\sigma}_{PTM}^{2} + \hat{\sigma}_{protein}^{2} \right) \right]
\leq \left( \frac{\Delta_{adj}}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2}
\end{equation}
where 
\begin{equation}
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1}
\end{equation}
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the Student distribution with $df = I(J-1)$ degrees of freedom.  Solving for $J$, the number of biological replicates per condition is
\begin{equation}
J \geq \frac{(2\hat{\sigma}_{PTM}^{2} + 2\hat{\sigma}_{protein}^{2})(t_{1-\beta, df} + t_{1-\alpha /2, df})^2}{\Delta_{adj}^2}
\end{equation}
Since the numerator has two sources of variation, reflecting a larger uncertainty in the adjusted calculation, the adjustment typically requires a larger sample size to gain the same sensitivity as the unadjusted estimation. Details of applications of this approach to experiments with complex design are in {\bf Supplementary Sec. \ref{supp-sec:complex_design}}.

@@

\todo{I suggest that you follow the same order as Figure 3, one paragraph per step. Describe each step, and refer to both the figure, the quantity in the equations and the MSstats functions. Specifically: design; data acquisition; data processing and requirements; converters; steps of statistical analysis. I would state upfront that all the extols before last are performed separately for each data type, and not repeat it in later paragraphs}

@@

$MSstatsPTM$ requires the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e. the modification site). This will generally include the amino acid abbreviation, plus it's number in the protein sequence. For example, a modification on a 70th amino acid in the sequence serine should be marked as ``S70''.

Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating where in the protein sequence the modification occurred. For these cases $MSstatsPTM$ includes functionality for identifying this location \todo{what do you mean by 'manually'? The users have to manually input it?} manually, given the modified peptide sequence and a FASTA file with the entire protein sequence. The package includes converters for seamless integration \todo{integration of what?} with multiple such tools. \todo{Maybe put all the discussion of converters in one place - integrate with the first paragraph? } The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites. If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect \todo{confound how? The effect of what? We use confounding to indicate the changes in protein abundance. This looks like a different confounding. Could you clarify?}. Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest \todo{confound how? The effect of what?}. For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.
